---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

First we need to unzip and load the data ...

```{r}
data <- read.csv(unz("activity.zip", "activity.csv"))

str(data)
```

In this analysis we'll use the dplyr, ggplot2 and gridExtra packages


```{r, message=FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)
```

## What is mean total number of steps taken per day?

To discover what's the mean total number os steps let's do:

1. Group the data by date, summarizing the steps with the sum function

```{r}
s1 <- data %>% 
        filter(!is.na(steps)) %>% 
        group_by(date) %>% 
        summarise(day_steps = sum(steps))
print(s1)
```

```{r}
summary(s1)
```

2. Print a histogram plot

```{r}
plot1 <- ggplot(s1, aes(x=day_steps)) + 
        geom_histogram(binwidth=3000, colour="black", fill="white") +
        geom_vline(xintercept=mean(s1$day_steps), color ="blue", linetype="dotted") +
        geom_vline(xintercept=median(s1$day_steps), color ="red", linetype="dashed") +
        labs(title="Steps per day", x="#Steps", y="Frecuency") +
        geom_text(aes(x=mean(s1$day_steps) + 2000, 
                      label="Mean  ", y=20), colour="blue", size=4 ) +
        geom_text(aes(x=median(s1$day_steps) + 2000, 
                      label="Median", y=19), colour="red", size=4)

plot1
```

The mean is **10766.19** and the median is **10765**

```{r}
mean(s1$day_steps)
median(s1$day_steps)
```

## What is the average daily activity pattern?

To discover what's the average daily activity pattern let's do:

1. Group the data by interval, summarizing the steps with the mean function

```{r}
s2 <- data %>% 
        filter(!is.na(steps)) %>% 
        group_by(interval) %>% 
        summarise(day_steps = mean(steps))
print(s2)
```

```{r}
summary(s2)
```

2. Find out which interval contains the maximum number of steps ...

```{r}
o <- s2 %>% 
        arrange(desc(day_steps)) 

max_interval <- as.numeric(o[1,1])
max_steps  <- as.integer(o[1,2])

print(max_interval)
print(max_steps)
```


3. Draw the line plot with the average pattern

```{r}


label_interval <- paste("Max interval = ", max_interval)
label_steps <-paste("Max steps = ", max_steps, "  ")
ggplot(s2, aes(x=interval, y=day_steps)) + 
        geom_line(colour="black", size=1) +
        geom_vline(xintercept=max_interval, color ="red", linetype="dashed") +
        labs(title="Daily pattern activity", x="Interval", 
             y="Steps") +
        geom_text(aes(x=max_interval + 300, y=200, label= label_interval), colour="red", size=4) +
        geom_text(aes(x=max_interval + 300, y=190, label= label_steps), colour="red", size=4)

```

The maximum number of steps is **208** which corresponds the the **835th** interval

## Imputing missing values

To calculate the total number of missing values in the dataset let's use the table function

```{r}
table(is.na(data$steps))
```

There are **2304** NA values in the dataset

The strategy for filling in the NA values will be fill with the mean value of all steps in the same interval.  To code this strategy let's do the following steps:

* First let's split the dataset into two diferents data frames, one with NA values and the other wihtout them.  

```{r}

fullData <- data[!is.na(data$steps),]
emptyData <- data[is.na(data$steps),]
```

* Calculate the mean steps in each interval


```{r}
m <- fullData %>% 
        group_by(interval) %>% 
        summarise(mean_steps = as.integer(mean(steps)))
```

* Fill in all the values of the empty data frame.

```{r}

for (row in 1:nrow(emptyData)){  
        row_interval <- emptyData[row, "interval"]
        mean_steps <- as.integer(
                m %>% 
                filter(interval == row_interval) %>%
                select(mean_steps)
        )
        
        emptyData[row, "steps"] <- mean_steps
}
```

* The last operation will be bind the two data frames (empty and full) into one named allData

```{r}
allData <- rbind(fullData, emptyData)

summary(allData)

```

* Draw an histogram plot with the two data frames at the same time

```{r}

s3 <- allData %>% 
        filter(!is.na(steps)) %>% 
        group_by(date) %>% 
        summarise(day_steps = sum(steps))

plot3 <- ggplot(s3, aes(x=day_steps)) + 
        geom_histogram(binwidth=3000, colour="black", fill="white") +
        geom_vline(xintercept=mean(s3$day_steps), color ="blue", linetype="dotted") +
        geom_vline(xintercept=median(s3$day_steps), color ="red", linetype="dashed") +
        labs(title="Steps per day [NA Filled]", x="#Steps", y="Frecuency") +
        geom_text(aes(x=mean(s3$day_steps) + 2000, 
                      label="Mean  ", y=20), colour="blue", size=4 ) +
        geom_text(aes(x=median(s3$day_steps) + 2000, 
                      label="Median", y=19), colour="red", size=4)

grid.arrange(plot1, plot3, ncol = 2, main = "Do differ the estimates when fill in missing data?")

  
```

* We can see that the median and the mean are almost equal for the two datasets, so when we use NA values **don't differ** from the estimates from the first part of the assignment

```{r}
summary(s1$day_steps)
```

```{r}
summary(s3$day_steps)
```

## Are there differences in activity patterns between weekdays and weekends?

First let's create a new factor column to divide between weekdays and weekends

```{r}
allData$date <- as.Date(allData$date)
allData$day_type <- as.factor(c("weekend", "weekday", "weekday", "weekday", "weekday", "weekday", "weekend")[as.POSIXlt(allData$date)$wday + 1])

table(allData$day_type)
```


Later let's group the data by interval and day type, summarizing the steps with the mean function

```{r}
s4 <- allData %>% 
        group_by(interval, day_type) %>% 
        summarise(day_steps = mean(steps))

print(s4)
```

```{r}
summary(s4)
```

Finally let draw a line plot coloring the data by day type so we can see the differences in activity patterns between weekdays and weekends.

```{r}

qplot(interval, day_steps, data = s4, color =day_type, geom = "line",
     xlab = "Interval", ylab = "Averaged steps") +  
        labs(title = "Weekday vs Weekend pattern activity")	

```

It looks like this person has a walking weekday pattern different from weekends.
